\section{Formalizing Ocaml Expressions}

Now we can begin to formalize expressions
in OCaml. We again re-iterate what steps are needed to build expressions
in our language, given that we have some \textit{context} now.

\begin{Def}[Building Expressions]

    When creating new expressions we must follow these steps:
    \begin{enumerate}
        \item \textbf{Context:} Define variable-to-type mappings.
        \item \textbf{Syntax:} Establish how the expression/operation should be written.
        \item \textbf{Typing Rules:} Define the type of the whole expression and its sub-expressions.
        \item \textbf{Semantics:} Clarify the resulting value/evaluation of the defined expression.
    \end{enumerate}
    I.e., what are our types, how are they used, what type of data do they represent, and how does it evaluate?
\end{Def}

\noindent
Now we begin to formalize, though we will abstract the context to $\Gamma$, assuming 
all the types we've defined before (\ref{tab:ocaml-types}).
\begin{Def}[Formalizing Let-Expressions]

    Let $\Gamma$ be the OCaml context, and $=$ be mathematical equality, and $\texttr{=}$ be an OCaml token:
    \begin{itemize}
        \item \textbf{Syntax:} \LARGE $\langle$\text{expr}$\rangle::=$ \texttr{let} $\langle$\text{var}$\rangle$ \texttr{=} $\langle$\text{expr}$\rangle$ \texttr{in} $\langle$\text{expr}$\rangle$ \normalsize\\
        
        \vspace{-.5em}
        \noindent
        If $x$ is a valid variable name, and $e_1$ is a well-formed expression and $e_2$ is a well-formed expression. Then $\texttr{let }x\texttr{ = }e_1\texttr{ in }e_2$ is also a well-formed expression.
        \item \textbf{Typing-Rule:} \LARGE $\dfrac{\Gamma\vdash e_1:\tau_1 \quad \Gamma,x:\tau_1\vdash e_2:\tau}{\Gamma\vdash\texttr{let }x\texttr{ = }e_1\texttr{ in }e_2:\tau}$ \normalsize\\
        
        \noindent
        Given context $\Gamma$, if there's some well-formed expression $e_1$ of type $\tau_1$ and some well-formed expression $e_2$ of type $\tau$, given a variable declaration of $x$ of type $\tau_1$, 
        then within this context, the expression $\texttr{let }x\texttr{  = }e_1\texttr{ in }e_2$ is of type $\tau$.
        
        \item \textbf{Semantics:} \LARGE $\dfrac{e_1\Downarrow v_1 \quad [v_1/x]e_2\Downarrow v}{\texttr{let }x\texttr{ = }e_1\texttr{ in }e_2\Downarrow v}$ \normalsize\\
        
        \noindent
        Following our context $\Gamma$, if a well-formed expression $e_1$ evaluates to $v_1$ and the substitution of $v_1$ for variable $x$ in another 
        well-formed expression $e_2$ evaluates to $v$, then the expression $\texttr{let }x\texttr{ = }e_1\texttr{ in }e_2$ evaluates to $v$.
    \end{itemize}

    \noindent
    Thus, we have formalized the \texttt{let} expression in OCaml.
\end{Def}

\newpage 

\noindent
Before we continue, we introduce the concept of $\top$ and $\bot$.

\begin{Def}[Top and Bottom ($\top$, $\bot$)]
    
    In logic and computer science:
    \begin{itemize}
        \item $\top$ is used to represent \textit{true}, \textit{valid}.
        \item $\bot$ is used to represent \textit{false} or \textit{invalid}.
    \end{itemize}

    \noindent
    Specifically, they are the greatest and least element of a lattice/boolean algebra (hence top and bottom), which when it comes to logic means truthhood and falsehood.
\end{Def}

\noindent
We continue with the formalization of the \texttt{if} expression in OCaml.
\begin{Def}[Formalizing If-Expressions]
    
    Let $\Gamma$ be the OCaml context, then:
    \begin{itemize}
        \item \textbf{Syntax:} \LARGE $\langle$\text{expr}$\rangle::=$ \texttr{if} $\langle$\text{expr}$\rangle$ \texttr{then} $\langle$\text{expr}$\rangle$ \texttr{else} $\langle$\text{expr}$\rangle$ \normalsize\\
     
        \vspace{-.5em}
        \noindent
        If $e_1$ is a well-formed expression, $e_2$ is a well-formed expression, and $e_3$ is a well-formed expression, then $\texttr{if }e_1\texttr{ then }e_2\texttr{ else }e_3$ is also a well-formed expression.

        \item \textbf{Typing-Rule:} \LARGE $\dfrac{\Gamma\vdash e_1:\texttt{bool} \quad \Gamma\vdash e_2:\tau \quad \Gamma\vdash e_3:\tau}{\Gamma\vdash\texttr{if }e_1\texttr{ then }e_2\texttr{ else }e_3:\tau}$ \normalsize\\

        \noindent
        Given context $\Gamma$, let there be well-formed expressions, $e_1$ of type \texttt{bool}, $e_2$ of type $\tau$, and $e_3$ of type $\tau$. Then the expression $\texttr{if }e_1\texttr{ then }e_2\texttr{ else }e_3$ is of type $\tau$.

        \item \textbf{Semantics:} \LARGE $\dfrac{e_1\Downarrow \top \quad e_2\Downarrow v}{\texttr{if }e_1\texttr{ then }e_2\texttr{ else }e_3\Downarrow v}$ (trueCond.) \normalsize\\
        
        \noindent
        Following our context $\Gamma$, if a well-formed expression $e_1$ evaluates $\top$ and another well-formed expression $e_2$ evaluates to $v$, then 
        the expression $\texttr{if }e_1\texttr{ then }e_2\texttr{ else }e_3$ evaluates to $v$ ($e_3$ a well-formed expression).
        \item \textbf{Semantics:} \LARGE $\dfrac{e_1\Downarrow \bot \quad e_3\Downarrow v}{\texttr{if }e_1\texttr{ then }e_2\texttr{ else }e_3\Downarrow v}$ (falseCond.) \normalsize\\
        
        \noindent
        Following our context $\Gamma$, if a well-formed expression $e_1$ evaluates $\bot$ and another well-formed expression $e_3$ evaluates to $v$, then
        the expression $\texttr{if }e_1\texttr{ then }e_2\texttr{ else }e_3$ evaluates to $v$ ($e_2$ a well-formed expression).
    \end{itemize}

    \noindent
    Take note that we must write two semantics rules for the \texttt{if} expression, one for when the condition evaluates to $\top$ and one for when it evaluates to $\bot$.
\end{Def}

\newpage 

\noindent
We continue with the formalization of the \texttt{function} expression in OCaml.
\begin{Def}[Formalizing Functions]

    \label{def:func}
    
    Let $\Gamma$ be the OCaml context, then:
    \begin{itemize}
        \item \textbf{Syntax:} \LARGE $\langle$expr$\rangle::=$\texttr{ fun }$\langle$var$\rangle$\texttr{ -> }$\langle$expr$\rangle$\normalsize\\

        \vspace{-.5em}
        \noindent
        If $x$ is a valid variable name and $e$ is a well-formed expression, then $\texttr{fun }x\texttr{ -> }e$ is also a well-formed expression.
        
        \item \textbf{Typing-Rule:} \LARGE $\dfrac{\Gamma,x:\tau_1\vdash e:\tau_2}{\Gamma\vdash\texttr{fun }x\texttr{ -> }e:\tau_1\rightarrow\tau_2}$ \normalsize
        
        \noindent
        Given context $\Gamma$ with a variable declaration of $(x:\tau_1)$ added, if there's a well-formed expression $e$ of type $\tau_2$ and, then the expression $\texttr{fun }x\texttr{ -> }e$ is of type $\tau_1\rightarrow\tau_2$.

        \item \textbf{Semantics:} \LARGE $\dfrac{}{\texttr{fun }x\texttr{ -> }e\Downarrow\lambda x.e}$ \normalsize\\
        
        \noindent
        Under no premises, the expression $\texttr{fun }x\texttr{ -> }e$ evaluates to the lambda function $\lambda x.e$.
    \end{itemize}
\end{Def}

\begin{Def}[Formalizing Application] 

    Let $\Gamma$ be the OCaml context, then:
    \begin{itemize}
        \item \textbf{Syntax:} \LARGE $\langle$expr$\rangle::=$\texttr{ }$\langle$expr$\rangle$\texttr{ }$\langle$expr$\rangle$\normalsize\\
        
        \vspace{-.5em}
        \noindent
        If $e_1$ is a well-formed expression and $e_2$ is a well-formed expression, then $e_1\ e_2$ is also a well-formed expression.

        \item \textbf{Typing-Rule:} \LARGE $\dfrac{\Gamma\vdash e_1:\tau_1\texttr{ -> }\tau_2 \quad \Gamma\vdash e_2:\tau_1}{\Gamma\vdash e_1\ e_2:\tau_2}$ \normalsize\\
        
        \noindent
        Given context $\Gamma$, if there's a well-formed expression $e_1$ of type $\tau_1$\texttr{ -> }$\tau_2$ (Functions.\ref{def:func}) and a well-formed expression $e_2$ of type $\tau_1$, then the expression $e_1\ e_2$ is of type $\tau_2$.

        \item \textbf{Semantics:} \LARGE $\dfrac{e_1\Downarrow\lambda x.e \quad e_2\Downarrow v \quad [v/x]e\Downarrow v'}{e_1\ e_2\Downarrow v'}$ \normalsize\\
        
        \noindent
        Following our context $\Gamma$, if a well-formed expression $e_1$ evaluates to a lambda function $\lambda x.e$, another well-formed expression $e_2$ evaluates to $v$, and the substitution of $v$ for $x$ in $e$ evaluates to $v'$, then the expression $e_1\ e_2$ evaluates to $v'$.
    \end{itemize}
\end{Def}

\newpage 

\noindent
Onto tuples and matching:

\begin{Def}[Formalizing Tuples]

    Let $\Gamma$ be the OCaml context, then:
    \begin{itemize}
        \item \textbf{Syntax:} \LARGE $\langle$expr$\rangle::=$\texttr{ (}$\langle$expr$\rangle$\texttr{,}$\dots$\texttr{,}$\langle$expr$\rangle$\texttr{)\normalsize}\normalsize\\
        
        \vspace{-.5em}
        \noindent
        If $e_1$ is a well-formed expression and $e_2$ is a well-formed expression, then $(e_1,e_2)$ is also a well-formed expression.

        \item \textbf{Typing-Rule:} \LARGE $\dfrac{\Gamma\vdash e_1:\tau_1 \quad \Gamma\vdash e_2:\tau_2 \quad \dots \quad \Gamma\vdash e_n:\tau_n}{\Gamma\vdash(e_1,e_2,\dots,e_n):\tau_1*\tau_2*\dots*\tau_n}$ \normalsize\\
        
        \noindent
        Given context $\Gamma$, if there are well-formed expressions $e_1$ of type $\tau_1$, $e_2$ of type $\tau_2$, and $e_n$ of type $\tau_n$, then the expression $(e_1,e_2,\dots,e_n)$ is of type $\tau_1*\tau_2*\dots*\tau_n$.

        \item \textbf{Semantics:} \LARGE $\dfrac{e_1\Downarrow v_1 \quad e_2\Downarrow v_2 \quad \dots \quad e_n\Downarrow v_n}{(e_1,e_2,\dots,e_n)\Downarrow(v_1,v_2,\dots,v_n)}$ \normalsize\\
        
        \noindent
        Following our context $\Gamma$, if well-formed expressions $e_1$ evaluates to $v_1$, $e_2$ evaluates to $v_2$, and $e_n$ evaluates to $v_n$, then the expression $(e_1,e_2,\dots,e_n)$ evaluates to $(v_1,v_2,\dots,v_n)$.
    \end{itemize}
\end{Def}

\begin{Def}[Formalizing Lists]

    Let $\Gamma$ be the OCaml context, then:
    \begin{itemize}
        \item \textbf{Syntax:} \LARGE 
        $\langle$expr$\rangle::=$ \texttr{[]} $\mid$ $\langle$expr$\rangle\ \texttr{::}\ \langle$expr$\rangle$ \normalsize\\
        
        \vspace{-.5em}
        \noindent
        The empty list \texttr{[]} is a well-formed expression. If $e_1$ is a well-formed expression and $e_2$ is a well-formed list, then $e_1\ \texttr{::}\ e_2$ is also a well-formed expression.

        \item \textbf{Typing-Rule:} \large \quad 
            $\dfrac{}{\Gamma\vdash []:\tau\ \texttr{list}}$ (nil)
            \quad
             $\dfrac{\Gamma\vdash e_1:\tau \quad \Gamma\vdash e_2:\tau\ \texttr{list}}{\Gamma\vdash e_1\ \texttr{::}\ e_2:\tau\ \texttr{list}}$ (cons) 
      

        \normalsize
        \noindent
        Given context $\Gamma$, the empty list \texttr{[]} has type $\tau\ \texttr{list}$ for any type $\tau$ (nil).
        If $e_1$ is of type $\tau$ and $e_2$ is of type $\tau\ \texttr{list}$, then the expression $e_1\ \texttr{::}\ e_2$ has type $\tau\ \texttr{list}$ (cons).

        \item \textbf{Semantics:} \large \quad
        $\dfrac{}{\texttr{[]} \Downarrow \varnothing}$ (nilEval)
        \quad $\dfrac{e_2 \Downarrow [v_2, ..., v_k] \quad e_1 \Downarrow v_1}{e_1\ \texttr{::}\ e_2 \Downarrow [v_1, v_2, ..., v_k]}$ (consEval) \normalsize\\
        
        \noindent
        The empty list \texttr{[]} evaluates to the empty list as a value (nilEval).
        If $e_2$ evaluates to the list $[v_2, ..., v_k]$ and $e_1$ evaluates to $v_1$, then $e_1\ \texttr{::}\ e_2$ evaluates to the list $[v_1, v_2, ..., v_k]$ (consEval).
    \end{itemize}
\end{Def}

\newpage

\noindent
Matching in a general sense is complex (deep matching), we can simplify with weak matching:

\begin{Def}[Weak Matching]

    Weak matching is a form of pattern matching that is for specific cases.
\end{Def}

\noindent
Additionally, we introduce the concept of \textit{side conditions} before we jump into weak matching on lists.

\begin{Def}[Side Conditions in Formal Semantics]

    A \textbf{side condition} is an additional constraint that must be satisfied before applying a rule. Side conditions are used to prevent undefined behavior and ensure correctness in evaluation.

    \textbf{Example 1: Integer Division Rule}

    \begin{itemize}
        \item Consider the evaluation rule for integer division:

        \LARGE
        \[
        \dfrac{e_1 \Downarrow v_1 \quad e_2 \Downarrow v_2 \quad v_2 \neq 0}
        {e_1 \div e_2 \Downarrow v_1 \div v_2}
        \]
        \normalsize

        \item The side condition \LARGE $v_2 \neq 0$ \normalsize ensures that:
        \begin{itemize}
            \item The denominator $v_2$ is not zero before performing division.
            \item If $v_2 = 0$, the rule cannot be applied to avoid division by zero.
        \end{itemize}

        \item Without this side condition, the expression could cause an error or undefined behavior.
    \end{itemize}

    \textbf{Example 2: Exponentiation with Non-Negative Exponents}

    \begin{itemize}
        \item Consider the evaluation rule for exponentiation:

        \LARGE
        \[
        \dfrac{e_1 \Downarrow v_1 \quad e_2 \Downarrow v_2 \quad v_2 \geq 0}
        {e_1^{e_2} \Downarrow v_1^{v_2}}
        \]
        \normalsize

        \item The side condition \LARGE $v_2 \geq 0$ \normalsize ensures that:
        \begin{itemize}
            \item The exponent $v_2$ is non-negative before applying the exponentiation operation.
            \item If $v_2 < 0$, the rule cannot be applied to avoid undefined results in integer arithmetic.
        \end{itemize}

        \item Without this side condition, expressions like $2^{-3}$ would be invalid in integer arithmetic.
    \end{itemize}

    Side conditions help enforce correctness by restricting operations to only valid inputs.

\end{Def}



\newpage

\begin{Def}[Weak Matching on Lists]

    Let $\Gamma$ be the OCaml context, then:
    \begin{itemize}
        \item \textbf{Syntax:} \LARGE 
        $\langle$expr$\rangle::=$ \texttr{match }$\langle$expr$\rangle$\texttr{ with }\\
        \text{}\hspace{7.5em}\texttr{| [] -> }$\langle$expr$\rangle$\\
        \text{}\hspace{7.5em}\texttr{| }$\langle$var$\rangle$ \texttr{:: }$\langle$var$\rangle$ \texttr{-> }$\langle$expr$\rangle$ \normalsize\\

        \vspace{-1em}
        \noindent
        If $e, e_1, e_2$ are well-formed expressions and $x, y$ are valid variable names, then\\
        $\texttr{match } e \texttr{ with | [] -> } e_1 \texttr{| } x \texttr{ :: } y \texttr{ -> } e_2$
        is a well-formed expression.

        \item \textbf{Typing Rule:}\\ \LARGE 
        
        $\dfrac{\Gamma\vdash e:\tau' \texttr{ list} \quad \Gamma\vdash e_1:\tau \quad \Gamma, x:\tau', y:\tau' \texttr{ list} \vdash e_2:\tau}
        {\Gamma\vdash \texttr{match } e \texttr{ with | [] -> } e_1 \texttr{| } x \texttr{ :: } y \texttr{ -> } e_2 : \tau}$ \normalsize\\
        
        \noindent
        If $e$ is of type $\tau' \texttr{ list}$ in the context $\Gamma$, and $e_1$ is of type $\tau$ in the context $\Gamma$, and $e_2$ is of type $\tau$ in the context $\Gamma$ with $(x:\tau')$ and $(y:\tau' \texttr{ list})$ added, then the entire match expression is of type $\tau$.

        \item \textbf{Semantics:}\\ 
        \LARGE 
            $\dfrac{e \Downarrow \varnothing \quad e_1 \Downarrow v}{\texttr{match } e \texttr{ with} \texttr{ [] -> } e_1 \texttr{| } x \texttr{:: } y \texttr{ -> } e_2 \Downarrow v}$ (nil) \normalsize\\
            
            \noindent
            If $e$ evaluates to the empty list $\varnothing$ and $e_1$ evaluates to $v$, then the entire match expression evaluates to $v$.

        \LARGE 
            $\dfrac{e \Downarrow h \texttr{ :: } t \quad e'_2 = [t/y][h/x]e_2 \quad e'_2 \Downarrow v}
            {\texttr{match } e \texttr{ with | [] -> } e_1 \texttr{| } x \texttr{ :: } y \texttr{ -> } e_2 \Downarrow v}$ (cons) \normalsize\\
            
            \noindent
            If $e$ evaluates to a nonempty list $h::t$ with first element $h$ and remainder $t$, and the expression $e_2$ with $h$ substituted for $x$ and $t$ substituted for $y$ evaluates to $v$, then the entire match expression evaluates to $v$.
       
    \end{itemize}
\end{Def}

\newpage

